---
title: "CyTOF workflow"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

Initialise packages
```{r}
#BiocManager::install(c("CATALYST","flowCore","cowplot","tidyverse","SingleCellExperiment","flowDensity","CytoNorm"))
packages <- c("CATALYST","flowCore","cowplot","tidyverse","SingleCellExperiment",
              "readxl","flowDensity","CytoNorm","cyCombine","diffcyt")
invisible(lapply(packages, library, character.only = TRUE))
```
Read in fcs files as flowCore flowSet
```{r}
donors <- c("07","09","12","13")
fcs_dirname <- "final_run_batches"
fcs_dir <- file.path(getwd(),fcs_dirname)
debar_gated_dir <- "final_run_norm_debar_fcs"
sample_key <- read.csv("CD45_12_plex_key.csv")
colnames(sample_key) <- as.numeric(c(89, 110, 111, 112, 114, 116))
rownames(sample_key) <- LETTERS[1:12]

for (i in seq_along(donors)){
donor <- donors[i]
fcs_data <- read.flowSet(path = fcs_dir,
                         pattern = paste0("*",donor,".fcs",sep=""),
                         transformation = NULL,
                         emptyValue = FALSE)
sce <- prepData(fcs_data)


####Normalise and remove beads with CATALYST

res <- normCytof(sce, beads = c(89,115,140,159,175,209), k = 50, 
  assays = c("counts", "exprs"), overwrite = FALSE)
n <- ncol(sce); ns <- c(ncol(res$beads), ncol(res$removed))
data.frame(
    check.names = FALSE, 
    "#" = c(ns[1], ns[2]), 
    "%" = 100*c(ns[1]/n, ns[2]/n),
    row.names = c("beads", "removed"))


####Extract normalised events

sce <- res$data
assayNames(sce)


####REFINE: Use flowAI and flowcut to clean flow rate and signal

#not sure if can be done in sce object


####Preliminary debarcoding
(sce <- assignPrelim(sce, sample_key))


####Refine debarcoding

sce <- estCutoffs(sce)
metadata(sce)$sep_cutoffs

sce <- applyCutoffs(sce, sep_cutoffs = min(metadata(sce)$sep_cutoffs))


####Convert barcoded data to flowSet

sce <- sce[, sce$bc_id != 0]
debar_fs <- sce2fcs(sce, split_by = "bc_id")


####Gate down to live singlets

fs_gated <- fsApply(debar_fs, intact_singlet_live_cells, ff_name = pData(debar_fs)$name)

#REFINE: montage all plots for QC purposes

####Save gated flow set and remove from workspace
write.flowSet(fs_gated, outdir = file.path(getwd(), debar_gated_dir, paste0("LC",donor,"_processed", sep = "")))
rm(fs_gated, sce)
}

```
Reorganise the outputted fcs files
```{r}
proj_dir <- "C:/Users/micha/OneDrive/Documents/R/Projects/2023_03___CyTOF"
indiv_donor_dirs <- list.dirs(path = debar_gated_dir, recursive = FALSE)
#rename files
for (i in seq_along(indiv_donor_dirs)){
  indiv_dir <- indiv_donor_dirs[i]
  setwd(file.path(proj_dir,indiv_dir))
  file.rename(list.files(file.path(proj_dir,indiv_dir), pattern="*.fcs"),
              gsub(".*fcs/","",
                   list.files(file.path(proj_dir,indiv_dir),
                              pattern = "*.fcs", full.names = TRUE)) %>% str_replace("processed/",""))
  setwd(proj_dir)
}
setwd(file.path(proj_dir))

#put all files into a single directory
file.copy(from=list.files(path = debar_gated_dir, pattern = "*.fcs", recursive = TRUE, full.names = TRUE),
          to="combined_batches")
```


Do cyCombine batch correction
```{r}
#library(cyCombine)
#
#batch_panel_file <- read_excel("panel_batch_combine.xlsx")
#batch_metadata_file <- read_excel("metadat_batch_combine.xlsx")
#
# Extract markers of interest
#markers <- batch_panel_file %>% 
#  filter(Type != "none") %>% 
#  pull(Antigen)

# Prepare a tibble from directory of FCS files
#uncorrected <- prepare_data(data_dir = "final_run_norm_debar_fcs",
#                             markers = markers,
#                             metadata = batch_metadata_file,
#                             panel = batch_panel_file,
#                             sample_ids = "Filename",
#                             batch_ids = "batch",
#                             filename_col = "Filename",
#                             condition = "condition",
#                             anchor = "anchor",
#                             panel_channel = "Channel",
#                             panel_antigen = "Antigen",
#                             down_sample = TRUE,
#                             sample_size = 500000,
#                             seed = 473,
#                             cofactor = 5) 
#
#detect_batch_effect_express(uncorrected, downsample = 10000, out_dir = 'batch_effect_check')
#
#corrected <- uncorrected %>%
#  batch_correct(covar = "condition",
#                xdim = 8,
#                ydim = 8,
#                norm_method = 'scale',
#                markers = markers)
```

Reload as sce object with metadata and panel info
```{r}
panel <- read_excel("panel.xlsx")
all_files_dir <- "combined_batches"

#change panel so downstream clustering not done on cisplatin or DNA

md <- read_excel("metadat_batch_combine.xlsx")
md$condition <- factor(md$condition, levels = c("a0_NT","a0_CAR","a1_NT","a1_CAR"))
md$sample_id <- factor(md$sample_id, 
    levels = md$sample_id[order(md$condition)])
sce <- prepData(file.path(proj_dir,all_files_dir), panel, md, features = panel$fcs_colname)
```

OR_load gated fcs files for CART data set
```{r}
panel <- read_excel("panel.xlsx")
gated_files_dir <- "new_gated_CAR_from_boolean"

#change panel so downstream clustering not done on cisplatin or DNA

md <- read_excel("metadat_batch_combine.xlsx")
md$condition <- factor(md$condition, levels = c("a0_NT","a0_CAR","a1_NT","a1_CAR"))
md$sample_id <- factor(md$sample_id, 
    levels = md$sample_id[order(md$condition)])
sce_gated <- prepData(file.path(proj_dir,gated_files_dir), panel, md, features = panel$fcs_colname)
```


Scan expression values per condition
```{r}
p <- plotExprs(sce_gated, color_by = "condition")
p$facet$params$ncol <- 6
```

Multi-dimensional scaling plot
```{r}
pbMDS(sce_gated, color_by = "condition", label_by = "sample_id")

```

Heatmap
```{r}
plotExprHeatmap(sce, scale = "last",
    hm_pal = rev(hcl.colors(10, "YlGnBu")), show_row_names = FALSE)
```

Do initial overclustering with FlowSOM and then recluster up to max 20 clusters with ConcensusClusterPLus - all in cluster() wrapper from CATALYST
```{r}
#Initial clustering with assigment of 100SOM clusters to up to 20 metaclusters
set.seed(1234)
sce <- cluster(sce, features = rownames(sce)[rownames(sce) %in% c("CD3", "CD4", "CD8a", "CD10", "CD19", "c-Myc")],
    xdim = 10, ydim = 10, maxK = 20, seed = 1234)

#Heatmap of expression per cluster
plotExprHeatmap(sce, features = rownames(sce)[rownames(sce) %in% c("CD3", "CD4", "CD8a", "CD10", "CD19", "c-Myc")], 
    by = "cluster_id", k = "meta5", 
    bars = TRUE, perc = TRUE)

#annotate 5 canonical cell clusters
meta5_annot <- c("B_cell", "CD4_NT", "CD4_CAR", "CD8_CAR", "CD8_NT")
merging_big_5 <- data.frame(c(1:5), factor(meta5_annot, levels = meta5_annot))
colnames(merging_big_5) <- c("original_cluster", "new_cluster")
sce <- mergeClusters(sce, k = "meta5", table = merging_big_5, id = "merged_canonical", overwrite = TRUE)

#Heatmap of expression per cluster after renaming
plotExprHeatmap(sce, features = rownames(sce)[rownames(sce) %in% c("CD3", "CD4", "CD8a", "CD10", "CD19", "c-Myc")], 
    by = "cluster_id", k = "merged_canonical", 
    bars = TRUE, perc = TRUE)

#Visualise with unap
set.seed(1234)
sce <- runDR(sce, "UMAP", cells = 1e3, features = rownames(sce_CART)[panel$marker_class != "none"])

#facet umaps by condition
plotDR(sce, "UMAP", color_by = "merged_canonical", facet_by = "condition")

```

Generate CAR-T only sce
```{r}
CART_dir <- "CART_fcs"
panel <- read_excel("panel.xlsx")

md_CART <- read_excel("CART_metadat_batch_combine.xlsx")
md_CART$condition <- factor(md_CART$condition, levels = c("a0_CAR","a1_CAR"))
md_CART$sample_id <- factor(md_CART$sample_id, 
    levels = md_CART$sample_id)
sce_CART <- prepData(file.path(proj_dir,CART_dir), panel, md_CART, features = panel$fcs_colname)
sce_CART <- sce_CART[,assay(sce_CART, "exprs")["c-Myc",]>1.875]
#recount numbers of cells
sample_id_names <- as.vector(metadata(sce_CART)$experiment_info$sample_id)
for(i in seq_along(sample_id_names)){
  metadata(sce_CART)$experiment_info$n_cells[i] <-
    sum(sce_CART$sample_id == as.vector(metadata(sce_CART)$experiment_info$sample_id[i]))
}

#p <- plotExprs(sce_CART, color_by = "condition")
#p$facet$params$ncol <- 6

#pbMDS(sce_CART, color_by = "condition", label_by = "sample_id")

```

Exploration CAR-T only data
```{r}
#sce_CART <- sce_gated
#Initial clustering with assigment of 100SOM clusters to up to 20 metaclusters
CART_diff_set <- c(5,6,7,9,10,12,13,17,19,21,24,27)
set.seed(1234)
sce_ctrl <- cluster(sce_ctrl, features = rownames(sce_CART)[CART_diff_set],
    xdim = 10, ydim = 10, maxK = 20, seed = 1234)

```

Visualisation CAR-T only data
```{r}
#Heatmap of expression per cluster
plotExprHeatmap(sce_ctrl, features = rownames(sce_CART)[CART_diff_set], 
    by = "cluster_id", k = "meta6", 
    bars = TRUE, perc = TRUE)

#Visualise with umap
set.seed(1234)
sce_gated <- runDR(sce_gated, "UMAP", cells = 1e3, features = rownames(sce_gated)[CART_diff_set])

#facet umaps by condition
plotDR(sce_ctrl, "UMAP", color_by = "meta6", facet_by = "condition")
```

Annotate clusters
```{r}
new_clusters <- c("CD8-EM", "CD8-EFF", "Noise", "CD4-EFF", "CD8-CM", "CD8-SCM-like", "CD4-EM", "CD4-CM-like")
meta8_annotate <- data.frame(c(1:8), factor(new_clusters)) %>%
  setNames(c("original_cluster","new_cluster"))

sce_ctrl <- mergeClusters(sce_ctrl, k = "meta8", table = meta8_annotate, id = "new_id")

```

Look at diff abundancies in CART only
```{r}
plotAbundances(sce_ctrl, k = "new_id", by = "cluster_id", shape_by = "patient_id")



```

DA of general protein expression in CART only
```{r}
sce_CART_overall_DE <- sce_CART
sce_CART_overall_DE <- mergeClusters(sce_CART_overall_DE, k = "meta20", id = "merging_all",
                                     table = data.frame(old_cluster = seq_len(20), new_cluster = "all"))

p <- plotMedExprs(sce_CART_overall_DE, features = rownames(sce_CART_overall_DE)[panel$marker_class != "none"],
                  shape_by = "patient_id")
p$facet$params$ncol <- 8

ei <- metadata(sce_CART_overall_DE)$experiment_info
ei$batch <- rep(c(1,2), each = 12)

design_mat <- createDesignMatrix(ei, cols_design = c("condition", "batch"))

contrast_mat <- createContrast(c(0, 1, 0)) # this contrast sepcifies preADI_CAR vs ctrl_CAR
block <- factor(ei$patient_id) # makes a pairwise comparison between individual donors 

rowData(sce_CART_overall_DE)$marker_class <- factor(c(rep("state",31), rep("none",4)))
ds_res_limma <- diffcyt(
  sce_CART_overall_DE,
  design = design_mat,
  contrast =  contrast_mat,
  analysis_type = "DS",
  clustering_to_use = "merging_all",
  verbose = TRUE,
  block_id = block,
)

#sample_reorder <- paste0(rep(c("LC07","LC09","LC12","LC13"), each = 6),
#                         rep("_", each = 24),
#                         rep(c(LETTERS[4:6],LETTERS[7:9]), times=4),
#                         sep="")

```

Plot single protein expression per cluster
```{r}
p <- plotMultiHeatmap(sce_CART, 
    hm1 = rownames(sce_CART)[CART_diff_set], hm2 = "TRAIL", k = "meta9", 
    row_anno = FALSE, bars = TRUE, perc = TRUE, normalize = FALSE)

```


Analyse differential protein expression between a0_CAR and a1_CAR
```{r}
#subset sce to include only control conditions
ctrl_conditions <- c("a0_CAR","a1_CAR")
sce_ctrl <- sce_gated[,colData(sce_gated)$condition %in% ctrl_conditions]
#filter metadata$experiment_info
metadata(sce_ctrl)$experiment_info <- metadata(sce_ctrl)$experiment_info[metadata(sce_ctrl)$experiment_info$condition %in%
                                                ctrl_conditions,]
#recount cells
sample_id_names_ctrl <- as.vector(metadata(sce_ctrl)$experiment_info$sample_id)
for(i in seq_along(sample_id_names)){
  metadata(sce_ctrl)$experiment_info$n_cells[i] <-
    sum(sce_ctrl$sample_id == as.vector(metadata(sce_ctrl)$experiment_info$sample_id[i]))
}

ei_ctrl <- metadata(sce_ctrl)$experiment_info %>%
  mutate_all(factor) #refactors all levels to remove filtered out conditions and samples
ei_ctrl$batch <- rep(c(1,2), each = 12)

design_mat_ctrl <- createDesignMatrix(ei_ctrl, cols_design = c("condition", "batch"))

contrast_mat_ctrl <- createContrast(c(0, 1, 0))
block_ctrl <- factor(ei_ctrl$patient_id) # makes a pairwise comparison between individual donors 

#reorder levels of sampl_ids to make HM look nice
sample_reorder <- paste0(rep(c("LC07","LC09","LC12","LC13"), each = 6),
                         rep("_", each = 24),
                         rep(c(LETTERS[4:6],LETTERS[7:9]), times=4),
                         sep="")
colData(sce_ctrl)$sample_id <- factor(colData(sce_ctrl)$sample_id, levels = sample_reorder)

ds_res_limma_ctrl <- diffcyt(
  sce_ctrl,
  design = design_mat_ctrl,
  contrast =  contrast_mat_ctrl,
  analysis_type = "DS",
  clustering_to_use = "new_id",
  verbose = TRUE,
  block_id = block_ctrl,
)

da_res_limma_ctrl <- diffcyt(
  sce_ctrl,
  design = design_mat_ctrl,
  contrast =  contrast_mat_ctrl,
  analysis_type = "DA",
  clustering_to_use = "new_id",
  verbose = TRUE,
  block_id = block_ctrl,
)


#rowData(ds_res_limma_ctrl$res)[rowData(ds_res_limma_ctrl$res)$ID == "1",] %>% data.frame() %>% arrange(p_adj, decreasing = FALSE)
# plotExprHeatmap(sce_ctrl, features = rownames(sce_CART)[CART_diff_set], 
#                 by = "cluster_id", k = "meta6", 
#                 bars = TRUE, perc = TRUE)

#plotDiffHeatmap(sce_ctrl, rowData(ds_res_limma_ctrl$res)[rowData(ds_res_limma_ctrl$res)$cluster_id %in% c("CD4-EFF", "CD4-EM", "CD4-CM-like") & rowData(ds_res_limma_ctrl$res)$p_adj < 0.05,], top_n = 50, fdr = FDR_cutoff, all = TRUE, sort_by = "lfc")
```



Per-cluster Differential analysis of sce
```{r}
FDR_cutoff <- 0.05

ei_sce <- metadata(sce)$experiment_info
ei_sce$preADI <- rep(c(0,1), each = 6, times = 4)
ei_sce$Tcell <- rep(c("NT", "CAR", "CAR", "NT"), each = 3, times = 4)
ei_sce$batch <- rep(c(0,1), each = 24)

design_mat_sce <- createDesignMatrix(ei_sce, cols_design = c("preADI", "Tcell", "batch"))
contrast_mat_sce <- createContrast(c(0, 1, 0, 0))
block_sce <- factor(ei_sce$patient_id)

marker_state_store <- rowData(sce)$marker_class
rowData(sce)$marker_class <- factor(c(rep("state",31), rep("none",4)))

sce_ds_res_limma <- diffcyt(
  sce,
  design = design_mat_sce,
  contrast =  contrast_mat_sce,
  analysis_type = "DS",
  clustering_to_use = "merged_canonical",
  verbose = TRUE,
  block_id = block_sce,
)

```

Volcano plot
```{r}
topTab <- rowData(ds_res_limma_ctrl$res)[!rowData(ds_res_limma_ctrl$res)$marker_id %in% c("CD10","CD19","c-Myc"),]
EV <- EnhancedVolcano::EnhancedVolcano(topTab,
                      x = "logFC",
                      y = "p_adj",
                      lab = topTab$marker_id,
                      xlim = c(-1.5, 0.5), ylim = c(0,22.5),
                      FCcutoff = 0,
                      legendPosition = '',
                      boxedLabels = TRUE,
                      caption = '',
                      col = c('grey','grey','grey','red3'),
                      drawConnectors = TRUE)





                



```



%%%%%%%%%%%%%%%%Functions from Computational and Structural Biotechnology Journal
Volume 19, 2021, Pages 3160-3175
Computational and Structural Biotechnology Journal
Database Article
Data processing workflow for large-scale immune monitoring studies by mass cytometry
Author links open overlay panelPaulina Rybakowska et al
%%%%%%%%%%%%%%%%
```{r}
#' gate_intact_cells
#' 
#' @description Performs gating of intact cells using flowDensity package
#'
#' @param flow_frame Character, full path to fcs_file.
#' @param file_name Character, the file name used only for plotting, if NULL
#' the file name stored in keyword GUID.original will be used, default is set to NULL
#' @param tinypeak_removal1 numeric from 0-1, as in deGate to exclude/include 
#' tiny peaks in the head of the density distribution curve for both Iridium 
#' channels
#' @param tinypeak_removal2 the same as tinypeak_removal1 but for the tail 
#' in the density distribution curve
#' @param alpha1 numeric, 0-1, as in deGate specify the significance of change 
#' in the slope being detected at the head of the density distribution curve
#' @param alpha2 the same as in alpha1 but for the tail of the density distribution curve
#' @param arcsine_transform Logical, if the data should be transformed 
#' with arcsine transformation and cofactor 5.
#' @param ... 
#' 
#' @return flow frame with intact cells

gate_intact_cells <- function(flow_frame, 
                              file_name = NULL,
                              tinypeak_removal1 = 0.8,
                              tinypeak_removal2 = 0.8,
                              alpha1 = 0.05,
                              alpha2 = 0.1, 
                              arcsine_transform = TRUE, ...){
  
  ff <- flow_frame
  
  if (is.null(file_name)){
    file_name <- ff_test
  } else {
    file_name 
  }
  
  if(arcsine_transform == TRUE){
    
    ff_t <- flowCore::transform(ff, 
                                flowCore::transformList(colnames(ff)[grep("Di", colnames(ff))], 
                                                        CytoNorm::cytofTransform))
  } else {
    ff_t <- ff
  }
  
  selection <- matrix(TRUE,
                      nrow = nrow(ff),
                      ncol = 1,
                      dimnames = list(NULL,
                                      c("intact")))
  
  tr <- list()
  for(m in c("Ir193Di", "Ir191Di")){
    
    tr[[m]] <- c(flowDensity::deGate(ff_t, m,
                                     tinypeak.removal = tinypeak_removal1, 
                                     upper = FALSE, use.upper = TRUE,
                                     alpha = alpha1, verbose = F, count.lim = 3, ...), 
                 flowDensity::deGate(ff_t, m,
                                     tinypeak.removal = tinypeak_removal2, 
                                     upper = TRUE, use.upper = TRUE,
                                     alpha = alpha2, verbose = F, count.lim = 3, ...)) 
  }
  
  for(m in c("Ir193Di", "Ir191Di")){
    selection[ff_t@exprs[,m] < tr[[m]][1], "intact"] <- FALSE
    selection[ff_t@exprs[,m] > tr[[m]][2], "intact"] <- FALSE
  }
  
  percentage <- (sum(selection)/length(selection))*100
  flowDensity::plotDens(ff_t, c("Ir193Di", "Ir191Di"))#, 
                        #main = paste0(file_name," ( ", format(round(percentage, 2), 
                         #                                               nsmall = 2), "% )"))
  
  abline(h = c(tr[["Ir191Di"]]))
  abline(v = c(tr[["Ir193Di"]]))
  points(ff_t@exprs[!selection[,"intact"], c("Ir193Di", "Ir191Di")], pch = ".")
  
  ff <- ff[selection[,"intact"], ]
  
  return(ff)
}


#' remove_mad_outliers
#'
#' @description detects outliers in the selected channel(s) using MAD 
#' (mean absolute deviation). 
#'
#' @param flow_frame 
#' @param channels character, channel names used for gating, default is set to 
#' "Event_length"
#' @param n_mad numeric, how many MAD should be use to detect outliers 
#' @param mad_f function used to compute deviation, default set to "mad" 
#' @param plot logicle, if to plot the data, default TRUE
#' @param center 
#' @param main character, title of the plot, default set to ""
#' @param ... other arguments to pass plotDens
#'
#' @return matrix with the selected cells 
remove_mad_outliers <- function(flow_frame, 
                                channels = "Event_length", 
                                n_mad = 2,
                                mad_f = mad,
                                plot = TRUE,
                                center = "center",
                                main = "",
                                ...){
  boundaries <- matrix(NA,
                       nrow = 5,
                       ncol = length(channels),
                       dimnames = list(c("median", "center", "mad", "l_lim", "u_lim"),
                                       channels))
  for (channel in channels) {
    x <- flow_frame@exprs[, channel]
    boundaries["median", channel] <- median(x)
    boundaries["center", channel] <- density(x)$x[which.max(density(x)$y)]
    boundaries["mad", channel] <- mad_f(x,
                                        center = boundaries[center, channel] )
    boundaries["l_lim", channel] <- boundaries[center, channel] - n_mad * boundaries["mad", channel]
    boundaries["u_lim", channel] <- boundaries[center, channel] + n_mad * boundaries["mad", channel]
  }
  
  selection <- rep(TRUE, nrow(flow_frame))
  for (channel in channels) {
    selection <- selection & (flow_frame@exprs[, channel] > boundaries["l_lim", channel])
    selection <- selection & (flow_frame@exprs[, channel] < boundaries["u_lim", channel])
  }
  percentage <- (sum(selection)/length(selection))*100
  if (plot) {
    flowDensity::plotDens(flow_frame, 
                          c(channels, "Ir191Di"))#, 
                          #main = paste0(main, " ( ", format(round(percentage, 2), 
                           #                                 nsmall = 2), "% )"),
                          #...)
    if(length(channels) == 2) {
      points(flow_frame@exprs[!selection, channels], col = "red", pch = ".")
      abline(v = boundaries[c("l_lim", "u_lim"), channels[1]], col = "grey")
      abline(h = boundaries[c("l_lim", "u_lim"), channels[2]], col = "grey")
    } else if(length(channels) == 1) {
      points(flow_frame@exprs[!selection, c(channels, "Ir191Di")], pch = ".")
      abline(v = boundaries[c("l_lim", "u_lim"), channels[1]], col = "grey")
    }
  }
  
  return(selection)
}


#' gate_singlet_cells
#'
#' @param flow_frame flow frame
#' @param channels character, channels name to be used for gating, default is 
#' to Event_length
#' @param arcsine_transform Logical, if the data should be transformed with 
#' arcsine transformation and cofactor 5.
#' @param file_name Character, the file name used only for plotting, if NULL
#' the file name stored in keyword GUID.original will be used, default is set to NULL
#' @param n_mad numeric, number of MADs to detect outliers
#' @param ... arguments to pass to plotDens
#'
#' @return flow frame with singlets 
gate_singlet_cells <- function(flow_frame, 
                               channels = "Event_length", 
                               arcsine_transform = TRUE,
                               file_name = NULL,
                               n_mad = 2,
                               ...){
  
  if (is.null(file_name)){
    file_name <- flow_frame
  } else {
    file_name 
  }
  
  if(arcsine_transform == TRUE){
    
    flow_frame_t <- flowCore::transform(flow_frame, 
                                        flowCore::transformList(colnames(flow_frame)[grep("Di", colnames(flow_frame))], 
                                                                CytoNorm::cytofTransform))
  } else {
    flow_frame_t <- flow_frame
  }
  
  selection <- matrix(TRUE,
                      nrow = nrow(flow_frame),
                      ncol = 1,
                      dimnames = list(NULL,
                                      c("singlets")))  
  
  selection[, "singlets"] <- remove_mad_outliers(flow_frame = flow_frame_t, 
                                                 channels = channels,
                                                 main = paste("Singlets", file_name),
                                                 n_mad = n_mad,
                                                 xlim = c(0, 100), ylim = c(0, 8), ...)
  
  flow_frame <- flow_frame[selection[,"singlets"], ]
  
  return(flow_frame)
  
}

#' gate_live_cells
#' 
#' @description Performs gating of live cells using flowDensity package
#' 
#' @param flow_frame Character, full path to fcs_file.
#' @param file_name Character, the file name used only for plotting, if NULL
#' the file name stored in keyword GUID.original will be used, default is set to NULL
#' @param viability_channel Character, the channel name used for viability staining
#' @param tinypeak_removal_viability, numeric from 0-1, as in deGate to exclude/include 
#' tiny peaks in the tail of the density ditribution curve for both viability channel 
#' @param tinypeak_removal_Iridium the same as tinypeak_removal_viablity but for
#' the head and tail of the density ditribution curve in Iridium channel
#' @param alpha_viability numeric, 0-1, as in deGate specify the significance of change 
#' in the slope of viability channel
#' @param alpha_Iridium the same as in alpha_viability but for the Iridium
#' @param arcsine_transform Logical, if the data should be transformed with 
#' arcsine transformation and cofactor 5.
#' @param ... arguments to pass to plotDens function
#' 
#' @return flow frame with live cells

gate_live_cells <- function(flow_frame, 
                            file_name = NULL,
                            viability_channel,
                            tinypeak_removal_viability = 0.8,
                            alpha_viability = 0.1,
                            tinypeak_removal_Iridium = 0.8,
                            alpha_Iridium = 0.05,
                            arcsine_transform = TRUE, ... ){
  
  ff <- flow_frame
  
  if (is.null(file_name)){
    file_name <- ff
  } else {
    file_name 
  }
  
  if(arcsine_transform == TRUE){
    
    ff_t <- flowCore::transform(ff, 
                                transformList(colnames(ff)[grep("Di", colnames(ff))], 
                                              CytoNorm::cytofTransform))
  } else {
    ff_t <- ff
  }
  
  selection <- matrix(TRUE,
                      nrow = nrow(ff),
                      ncol = 1,
                      dimnames = list(NULL,
                                      c("live")))
  
  
  v_ch <- grep(viability_channel, colnames(ff), value = T)
  
  tr <- list()
  for(m in c("Ir191Di", v_ch)){
    if (m == v_ch) {
      upper = FALSE
      alpha = alpha_viability
      tr[[m]] <- flowDensity::deGate(ff_t, m,
                                     #tinypeak.removal = tinypeak_removal_viability, 
                                     #upper = upper, use.upper = FALSE,
                                     #alpha = alpha, verbose = F, count.lim = 3,
                                     bimodal = TRUE)
      
    } else {
      alpha = alpha_Iridium
      tr[[m]] <- c(flowDensity::deGate(ff_t, m,
                                       tinypeak.removal = tinypeak_removal_Iridium, 
                                       upper = FALSE, use.upper = TRUE,
                                       alpha = alpha,  verbose = F, count.lim = 3), 
                   flowDensity::deGate(ff_t, m,
                                       tinypeak.removal = tinypeak_removal_Iridium, 
                                       upper = TRUE, use.upper = TRUE,
                                       alpha = alpha, verbose = F, count.lim = 3)) 
      
    }
  }
  
  for(m in c(v_ch, "Ir191Di")){
    if (m == v_ch) {
      selection[ff_t@exprs[,m] > tr[[m]][1], "live"] <- FALSE 
    } else {
      selection[ff_t@exprs[,m] < tr[[m]][1], "live"] <- FALSE
      selection[ff_t@exprs[,m] > tr[[m]][2], "live"] <- FALSE  
    }
  }
  percentage <- (sum(selection)/length(selection))*100
  flowDensity::plotDens(ff_t, c(v_ch, "Ir191Di"), 
                        #main = paste0(file_name," ( ", format(round(percentage, 2), nsmall = 2), "% )"),
                        xlim = c(0, 8), ylim = c(0, 8), ...)
  
  abline(h = tr[["Ir191Di"]])
  abline(v = tr[[v_ch]])
  
  points(ff_t@exprs[!selection[,"live"], c(v_ch, "Ir191Di")], pch = ".") 
  
  ff <- ff[selection[,"live"], ]
  
  return(ff)
  
}

intact_singlet_live_cells <- function(ff, ff_name){  
  #dev.off()
  #par(mfrow = c(12,3))
  ff <- gate_intact_cells(flow_frame = ff, 
                          file_name = ff_name)
  
  ff <- gate_singlet_cells(flow_frame = ff,
                           channels = "Event_length",
                           file_name = ff_name)
  
  ff <- gate_live_cells(flow_frame = ff, 
                        viability_channel = "Pt195Di",
                        file_name = ff_name)
}

```